---
title: "TROST Select: LASSO Model"
author: "Heike Sprenger"
date: "07/06/2020"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_float: true
    toc_depth: 4
editor_options: 
  chunk_output_type: console
---

# Setup
```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
# opts_knit$set(root.dir = '../')

library(caret)
library(glmnet)
library(tidyverse)
library(pcaMethods)

# load workspace
# load("lasso_model.RData")
```

# Load functions
```{r load functions}
source("func_prep_pca.R")
source("func_remove_factors.R")
```


# Load data
## Tolerance information
```{r tolerance information}
tolerance <- read.table("tolerance.txt", sep = "\t", header = T)
```


## Transcript data

**TROST: 202 samples, 42 transcripts**

**all data: 1159 samples**

```{r load transcript data}
# ALL
transcript_data_all <- read.table("log_norm_ct_trost_valdis.txt", header = T, sep = "\t")
transcript_samples_all <- 
  read.table("transcript_samplelist_trost_valdis.txt", header=TRUE, sep="\t") %>%
  left_join(tolerance[,-3], by = "subspecies_id")

transcript_samples_all$row_id <- rownames(transcript_data_all)

# VALDIS lines
transcript_samples_valdis <- 
  read.table("transcript_samplelist_valdis.txt", header = T, sep = "\t") %>%
  dplyr::select(line_id, crossing, name) %>% distinct
```

## Metabolite data

**911 samples, XXX metabolites**

```{r load metabolite data}
metabolite_data_all <- read.table("metabolite_data_trost_valdis.txt", 
                                  header = T, sep = "\t", check.names = F)
dim(metabolite_data_all) 
# 2336 samples, 81 metabolites

metabolite_samples_all <- 
  read.table("metabolite_samplelist_trost_valdis.txt", header=TRUE, sep="\t") %>%
  left_join(tolerance, by = "cultivar")

# Import analytes overlap table (overlap regarding 17 measured TROST experiments, not the QC experiments) includes 81 analytes
analytes <- 
  read.table("analytes_trost_valdis.txt", sep = "\t", header = T) %>% 
  filter(select_part == "yes") %>%
  arrange(analyte) %>% droplevels

length(analytes$name) # 81
colnames(metabolite_data_all) <- analytes$MPIMP_ID
```

# Remove batch effect in metabolite data
```{r Remove batch effect in metabolite data}
metabolite_data_norm <- apply(metabolite_data_all, 2, 
                              RemoveFactors, 
                              sam = metabolite_samples_all, 
                              facs=c("genotype", "treatment", "SequenceID", "BatchID", "log10_AvgAnnotated"), 
                              keep=c("genotype", "treatment"))

sum(is.na(metabolite_data_norm)) / (ncol(metabolite_data_norm)*nrow(metabolite_data_norm))*100
# 4.5%
```


# PCA
## Transcripts
```{r PCA transcripts}
pca_transcript_all <- func_prep_pca(transcript_data_all, 
                                     scale_method = "none", 
                                     center_option = FALSE, 
                                     pc_number = 5, 
                                     pca_method = "rnipals")

# complete observations
compObs_transcript_all <- 
  pca_transcript_all@completeObs %>%
  as.data.frame
```

## Metabolites
```{r PCA metabolites}
pca_metabolite_all <- func_prep_pca(metabolite_data_norm, 
                                    scale_method = "none", 
                                    center_option = FALSE, 
                                    pc_number = 10, 
                                    pca_method = "rnipals")

# complete observations
compObs_metabolite_all <- 
  pca_metabolite_all@completeObs %>%
  as.data.frame
```


# Define training and test data

**Subsets: TROST and VALDIS**

## Transcripts

### TROST data for training model

```{r TROST subset transcripts}
transcript_samples_training <- 
  transcript_samples_all %>%
  filter(cultivation == "field") %>%
  filter(trost_valdis == "trost") %>%
  filter(!is.na(model_set)) %>%
  droplevels()
  
compObs_transcript_training <-
  compObs_transcript_all %>%
  rownames_to_column("row_id") %>%
  filter(row_id %in% transcript_samples_training$row_id) %>%
  column_to_rownames("row_id")

dim(compObs_transcript_training) # 202 samples
```

### VALDIS data for model prediction

```{r VALDIS subset transcripts}
transcript_samples_pred <- 
  transcript_samples_all %>%
  filter(trost_valdis == "valdis") %>%
  droplevels()

compObs_transcript_pred <-
  compObs_transcript_all %>%
  rownames_to_column("row_id") %>%
  filter(row_id %in% transcript_samples_pred$row_id) %>%
  column_to_rownames("row_id")

dim(compObs_transcript_pred) # 803 samples
```

## Metabolites

### TROST data for training model

```{r TROST subset metabolites}
metabolite_samples_training <- 
  metabolite_samples_all %>%
  filter(cultivation == "field") %>%
  filter(trost_valdis == "trost") %>%
  filter(!is.na(model_set)) %>%
  droplevels()

compObs_metabolite_training <-
  compObs_metabolite_all %>%
  rownames_to_column("chromatogram") %>%
  filter(chromatogram %in% metabolite_samples_training$chromatogram) %>%
  column_to_rownames("chromatogram")

dim(compObs_metabolite_training) # 911 samples
```

### VALDIS transcript data for model prediction

```{r VALDIS subset metabolites}
metabolite_samples_pred <- 
  metabolite_samples_all %>%
  filter(trost_valdis == "valdis") %>%
  droplevels()

compObs_metabolite_pred <-
  compObs_metabolite_all %>%
  rownames_to_column("chromatogram") %>%
  filter(chromatogram %in% metabolite_samples_pred$chromatogram) %>%
  column_to_rownames("chromatogram")

dim(compObs_metabolite_pred) # 806 samples
```

# LASSO Model
## Transcripts
### Moodel training

``mdrym_fve``: Median DRYM

```{r LASSO model training transcripts}
set.seed(1)
lasso_fit_transcript <- glmnet(x = as.matrix(compObs_transcript_training), 
                                y = transcript_samples_training$mdrym_fve)
plot(lasso_fit_transcript)

# Cross-validation (10-fold)
set.seed(1111)
lasso_cv_transcript <- glmnet::cv.glmnet(x = as.matrix(compObs_transcript_training), 
                                 y = transcript_samples_training$mdrym_fve)
lasso_cv_transcript$lambda.1se # 0.001209806
plot(lasso_cv_transcript, ylim=c(0, 0.005))

lasso_cv_transcript_coef_1se <- predict(lasso_cv_transcript, 
                                         type = "coefficients", 
                                         s = lasso_cv_transcript$lambda.1se)
table(as.matrix(lasso_cv_transcript_coef_1se) == 0) # 23 transcripts left
```

### Predict DRYM for VALDIS data
```{r Predict DRYM for VALDIS transcript data}
# use largest value of lambda such that CV-error is within 1 standard error of the minimum (ca. 50 metabolites)
lasso_transcript_predicted_drym_valdis_1se <- 
  predict(lasso_cv_transcript, 
          new = as.matrix(compObs_transcript_pred), 
          s = "lambda.1se")

lasso_transcript_predicted_drym_valdis <- 
  predict(lasso_cv_transcript, 
          new = as.matrix(compObs_transcript_pred), 
          s = "lambda.min")
```

### Median DRYM values
```{r lasso sort and aggregate drym values transcripts}
# join predicted DRYM values with line IDs and calculate median
lasso_transcript_predicted_drym_valdis_1se_median <- 
  data.frame(drym = lasso_transcript_predicted_drym_valdis_1se[,1], 
             line_id = transcript_samples_pred$subspecies_id) %>%
  group_by(line_id) %>%
  summarize(median_drym = median(drym)) %>%
  rename(drym = median_drym) %>%
  left_join(transcript_samples_valdis, by = "line_id")

lasso_transcript_predicted_drym_valdis_median <- 
  data.frame(drym = lasso_transcript_predicted_drym_valdis[,1], 
             line_id = transcript_samples_pred$subspecies_id) %>%
  group_by(line_id) %>%
  summarize(median_drym = median(drym)) %>%
  rename(drym = median_drym) %>%
  left_join(transcript_samples_valdis, by = "line_id")

# save median of predicted drym
write.table(lasso_transcript_predicted_drym_valdis_1se_median, 
            "lasso_transcript_predicted_drym_valdis_1se_median.txt", 
            sep = "\t", row.names = F)
```

## Metabolites
### Moodel training

``mdrym_fve``: Median DRYM

```{r LASSO model training metabolites}
set.seed(1)
lasso_fit_metabolite <- glmnet(x = as.matrix(compObs_metabolite_training), 
                                y = metabolite_samples_training$mdrym_fve)
plot(lasso_fit_metabolite)

# Cross-validation (10-fold)
set.seed(1111)
lasso_cv_metabolite <- glmnet::cv.glmnet(x = as.matrix(compObs_metabolite_training), 
                                 y = metabolite_samples_training$mdrym_fve)
lasso_cv_metabolite$lambda.1se # 0.0006964521
plot(lasso_cv_metabolite, ylim=c(0, 0.005))

# define lambda for sparse model with 29 variables
lambda.sparse <- 0.002

lasso_cv_metabolite_coef_1se <- predict(lasso_cv_metabolite, 
                                         type = "coefficients", 
                                         s = lambda.sparse)
table(as.matrix(lasso_cv_metabolite_coef_1se) == 0)
```

### Predict DRYM for VALDIS data
```{r Predict DRYM for VALDIS metabolite data}
# use largest value of lambda such that CV-error is within 1 standard error of the minimum (ca. 50 metabolites)
lasso_metabolite_predicted_drym_valdis_sparse <- 
  predict(lasso_cv_metabolite, 
          new = as.matrix(compObs_metabolite_pred), 
          s = lambda.sparse)

lasso_metabolite_predicted_drym_valdis <- 
  predict(lasso_cv_metabolite, 
          new = as.matrix(compObs_metabolite_pred), 
          s = "lambda.min")
```

### Median DRYM values
```{r lasso sort and aggregate drym values metabolites}
# join predicted DRYM values with line IDs and calculate median
lasso_metabolite_predicted_drym_valdis_sparse_median <- 
  data.frame(drym = lasso_metabolite_predicted_drym_valdis_sparse[,1], 
             line_id = metabolite_samples_pred$cultivar) %>%
  group_by(line_id) %>%
  summarize(median_drym = median(drym)) %>%
  rename(drym = median_drym) %>%
  mutate(name = str_replace(line_id, "_", "")) %>%
  mutate(name = str_replace(name, "AxR", "AR")) %>%
  mutate(name = str_replace(name, "ExA", "EA")) %>%
  mutate(name = str_replace(name, "ALBATROS", "Albatros")) %>%
  mutate(name = str_replace(name, "DESIREE", "Desiree")) %>%
  mutate(name = str_replace(name, "EURORESA", "Euroresa")) %>%
  mutate(name = str_replace(name, "RAMSES", "Ramses"))

lasso_metabolite_predicted_drym_valdis_median <- 
  data.frame(drym = lasso_metabolite_predicted_drym_valdis[,1], 
             line_id = metabolite_samples_pred$cultivar) %>%
  group_by(line_id) %>%
  summarize(median_drym = median(drym)) %>%
  rename(drym = median_drym) %>%
  mutate(name = str_replace(line_id, "_", "")) %>%
  mutate(name = str_replace(name, "AxR", "AR")) %>%
  mutate(name = str_replace(name, "ExA", "EA")) %>%
  mutate(name = str_replace(name, "ALBATROS", "Albatros")) %>%
  mutate(name = str_replace(name, "DESIREE", "Desiree")) %>%
  mutate(name = str_replace(name, "EURORESA", "Euroresa")) %>%
  mutate(name = str_replace(name, "RAMSES", "Ramses"))

# save median of predicted drym
write.table(lasso_metabolite_predicted_drym_valdis_sparse_median, 
            "lasso_metabolite_predicted_drym_valdis_sparse_median.txt", 
            sep = "\t", row.names = F)
```

## Combine predicted DRYM values
```{r Combine predicted DRYM values}
lasso_predicted_drym_sparse <-
  lasso_metabolite_predicted_drym_valdis_sparse_median %>%
  dplyr::select(name, drym) %>%
  rename(drym_metabolite = drym) %>%
  left_join(lasso_transcript_predicted_drym_valdis_1se_median, by = "name") %>%
  rename(drym_transcript = drym) %>%
  mutate(drym_avg = (drym_metabolite + drym_transcript)/2) %>%
  arrange(drym_avg)

lasso_predicted_drym <-
  lasso_metabolite_predicted_drym_valdis_median %>%
  dplyr::select(name, drym) %>%
  rename(drym_metabolite = drym) %>%
  left_join(lasso_transcript_predicted_drym_valdis_median, by = "name") %>%
  rename(drym_transcript = drym) %>%
  mutate(drym_avg = (drym_metabolite + drym_transcript)/2) %>%
  arrange(drym_avg)
```

## Subpopulations
```{r Subpopulations}
lines_MPt <- c("AR1", "AR23", "AR56", "AR67", "AR106", "AR121", "AR157", "AR163", "AR183", 
               "AR185", "AR196", "AR197", "AR200", "AR241", "AR245", "AR254", "AR269", "AR282",
               "AR285", "AR293", "EA28", "EA74", "EA87")
lines_MPs <- c("AR55", "EA2", "EA8", "EA19", "EA22", "EA54", "EA55", "EA71", "EA92", 
               "EA111", "EA112", "EA131", "EA154", "EA165", "EA172", "EA173", "EA174", 
               "EA252", "EA269", "EA273", "EA279", "EA280")
  
lasso_predicted_drym_sparse$select <- "not"
lasso_predicted_drym_sparse$select[which(lasso_predicted_drym_sparse$name %in% lines_MPt)] <- "MPt"
lasso_predicted_drym_sparse$select[which(lasso_predicted_drym_sparse$name %in% lines_MPs)] <- "MPs"
table(lasso_predicted_drym_sparse$select)

lasso_predicted_drym$select <- "not"
lasso_predicted_drym$select[which(lasso_predicted_drym$name %in% lines_MPt)] <- "MPt"
lasso_predicted_drym$select[which(lasso_predicted_drym$name %in% lines_MPs)] <- "MPs"
table(lasso_predicted_drym$select)

ggplot(lasso_predicted_drym_sparse, aes(x = select, y = drym_avg)) + 
  geom_boxplot()
```


# Plots
## Transcripts
### Boxplot of predicted DRYM vs line
```{r boxplot of predicted DRYM vs line transcripts}
lasso_transcript_predicted_drym_valdis_df <- 
  data.frame(drym = lasso_transcript_predicted_drym_valdis[,1], 
             line_id = transcript_samples_pred$subspecies_id) %>%
  left_join(transcript_samples_valdis, by = "line_id")

# sort predicted DRYM values for plot
lasso_transcript_predicted_drym_valdis_sorted <- 
  with(lasso_transcript_predicted_drym_valdis_df, reorder(name, drym, median, na.rm=T))

lasso_color <- rep("#00756D", 199) # for ExA
lasso_color [which( grepl("^AR", levels(lasso_transcript_predicted_drym_valdis_sorted) ))] <- "#BF5300"
lasso_color [which( levels (lasso_transcript_predicted_drym_valdis_sorted) == "Albatros" )] <- "grey"
lasso_color [which( levels (lasso_transcript_predicted_drym_valdis_sorted) == "Euroresa" )] <- "#5778B9"
lasso_color [which( levels (lasso_transcript_predicted_drym_valdis_sorted) == "Ramses" )] <- "#F7B944"
lasso_color [which( levels (lasso_transcript_predicted_drym_valdis_sorted) == "Desiree" )] <- "white"
# lasso_color <- subset(lasso_color, levels (lasso_transcript_predicted_drym_valdis_sorted) != "DESIREE")
lasso_color_fac <- factor(lasso_color)

png("boxplot_lasso_transcript_predicted_drym_valdis_full.png", width=3000, height=1500, res=300)
par(mar=c(4.5,5,2,0.5))
boxplot(drym ~ lasso_transcript_predicted_drym_valdis_sorted, data = lasso_transcript_predicted_drym_valdis_df, 
        ylab="DRYM", cex.lab=1.5, cex.axis=1.2, las=2, col=lasso_color, main = "transcript model", names=F, xlab="genotype")
legend("bottomright", fill=levels(lasso_color_fac), legend=c("ExA", "E", "AxR", "R", "A"), horiz=T)
dev.off()
```

### Boxplot of predicted DRYM vs line (sparse)
```{r boxplot of predicted DRYM vs line transcripts sparse}
lasso_transcript_predicted_drym_valdis_1se_df <- 
  data.frame(drym = lasso_transcript_predicted_drym_valdis_1se[,1], 
             line_id = transcript_samples_pred$subspecies_id) %>%
  left_join(transcript_samples_valdis, by = "line_id")

# sort predicted DRYM values for plot
lasso_transcript_predicted_drym_valdis_1se_sorted <- 
  with(lasso_transcript_predicted_drym_valdis_1se_df, reorder(name, drym, median, na.rm=T))

lasso_color <- rep("#00756D", 199) # for ExA
lasso_color [which( grepl("^AR", levels(lasso_transcript_predicted_drym_valdis_1se_sorted) ))] <- "#BF5300"
lasso_color [which( levels (lasso_transcript_predicted_drym_valdis_1se_sorted) == "Albatros" )] <- "grey"
lasso_color [which( levels (lasso_transcript_predicted_drym_valdis_1se_sorted) == "Euroresa" )] <- "#5778B9"
lasso_color [which( levels (lasso_transcript_predicted_drym_valdis_1se_sorted) == "Ramses" )] <- "#F7B944"
lasso_color [which( levels (lasso_transcript_predicted_drym_valdis_1se_sorted) == "Desiree" )] <- "white"
# lasso_color <- subset(lasso_color, levels (lasso_transcript_predicted_drym_valdis_1se_sorted) != "DESIREE")
lasso_color_fac <- factor(lasso_color)

png("boxplot_lasso_transcript_predicted_drym_valdis_1se.png", width=3000, height=1500, res=300)
par(mar=c(4.5,5,2,0.5))
boxplot(drym ~ lasso_transcript_predicted_drym_valdis_1se_sorted, data = lasso_transcript_predicted_drym_valdis_1se_df, 
        ylab="DRYM", cex.lab=1.5, cex.axis=1.2, las=2, col=lasso_color, main = "transcript model", names=F, xlab="genotype")
legend("bottomright", fill=levels(lasso_color_fac), legend=c("ExA", "E", "AxR", "R", "A"), horiz=T)
dev.off()
```


## Metabolites
### Boxplot of predicted DRYM vs line
```{r boxplot of predicted DRYM vs line metabolites}
lasso_metabolite_predicted_drym_valdis_df <- 
  data.frame(drym = lasso_metabolite_predicted_drym_valdis[,1], 
             line_id = metabolite_samples_pred$cultivar)

# sort predicted DRYM values for plot
lasso_metabolite_predicted_drym_valdis_sorted <- 
  with(lasso_metabolite_predicted_drym_valdis_df, reorder(line_id, drym, median, na.rm=T))

lasso_color <- rep("#00756D", 199) # for ExA
lasso_color [which( grepl("^AxR", levels(lasso_metabolite_predicted_drym_valdis_sorted) ))] <- "#BF5300"
lasso_color [which( levels (lasso_metabolite_predicted_drym_valdis_sorted) == "ALBATROS" )] <- "grey"
lasso_color [which( levels (lasso_metabolite_predicted_drym_valdis_sorted) == "EURORESA" )] <- "#5778B9"
lasso_color [which( levels (lasso_metabolite_predicted_drym_valdis_sorted) == "RAMSES" )] <- "#F7B944"
lasso_color [which( levels (lasso_metabolite_predicted_drym_valdis_sorted) == "DESIREE" )] <- "white"
# lasso_color <- subset(lasso_color, levels (lasso_metabolite_predicted_drym_valdis_sorted) != "DESIREE")
lasso_color_fac <- factor(lasso_color)

png("boxplot_lasso_metabolite_predicted_drym_valdis_full.png", width=3000, height=1500, res=300)
par(mar=c(4.5,5,2,0.5))
boxplot(drym ~ lasso_metabolite_predicted_drym_valdis_sorted, data = lasso_metabolite_predicted_drym_valdis_df, 
        ylab="DRYM", cex.lab=1.5, cex.axis=1.2, las=2, col=lasso_color, main = "metabolite model", names=F, xlab="genotype")
legend("bottomright", fill=levels(lasso_color_fac), legend=c("ExA", "E", "AxR", "R", "A"), horiz=T)
dev.off()
```

### Boxplot of predicted DRYM vs line (sparse)
```{r boxplot of predicted DRYM vs line metabolites sparse}
lasso_metabolite_predicted_drym_valdis_sparse_df <- 
  data.frame(drym = lasso_metabolite_predicted_drym_valdis_sparse[,1], 
             line_id = metabolite_samples_pred$cultivar)

# sort predicted DRYM values for plot
lasso_metabolite_predicted_drym_valdis_sparse_sorted <- 
  with(lasso_metabolite_predicted_drym_valdis_sparse_df, reorder(line_id, drym, median, na.rm=T))

lasso_color <- rep("#00756D", 199) # for ExA
lasso_color [which( grepl("^AxR", levels(lasso_metabolite_predicted_drym_valdis_sparse_sorted) ))] <- "#BF5300"
lasso_color [which( levels (lasso_metabolite_predicted_drym_valdis_sparse_sorted) == "ALBATROS" )] <- "grey"
lasso_color [which( levels (lasso_metabolite_predicted_drym_valdis_sparse_sorted) == "EURORESA" )] <- "#5778B9"
lasso_color [which( levels (lasso_metabolite_predicted_drym_valdis_sparse_sorted) == "RAMSES" )] <- "#F7B944"
lasso_color [which( levels (lasso_metabolite_predicted_drym_valdis_sparse_sorted) == "DESIREE" )] <- "white"
# lasso_color <- subset(lasso_color, levels (lasso_metabolite_predicted_drym_valdis_sparse_sorted) != "DESIREE")
lasso_color_fac <- factor(lasso_color)

png("boxplot_lasso_metabolite_predicted_drym_valdis_sparse.png", width=3000, height=1500, res=300)
par(mar=c(4.5,5,2,0.5))
boxplot(drym ~ lasso_metabolite_predicted_drym_valdis_sparse_sorted, data = lasso_metabolite_predicted_drym_valdis_sparse_df, 
        ylab="DRYM", cex.lab=1.5, cex.axis=1.2, las=2, col=lasso_color, main = "metabolite model", names=F, xlab="genotype")
legend("bottomright", fill=levels(lasso_color_fac), legend=c("ExA", "E", "AxR", "R", "A"), horiz=T)
dev.off()
```


# Save workspace
```{r save workspace}
save.image("lasso_model.RData")
```


# Session Info
```{r session info}
sessionInfo()
```

